Longitudinal Principal Manifold Estimation

### 2024-02-12 paper/main.tex
update application, discussion sections, figures

\usepackage[
backend=biber,
style=authoryear
]{biblatex}

@@

\begin{figure}[ht]
  \centering
  \subfloat[\centering Data]{{\includegraphics[width=8cm]{sim_case1_plot2}}}%
  \hfill
  \subfloat[\centering Step 1: Sample Size Reduction]{{\includegraphics[width=8cm]{lpme_data_reduction}}}
  \vfill
  \subfloat[\centering Step 2: Initialization]{{\includegraphics[width=8cm]{lpme_initialization}}}
  \hfill
  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\includegraphics[width=8cm]{lpme_fitting}}}
  \caption{LPME Algorithm Steps. In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface. }
  \label{fig:lpme_steps}
\end{figure}

@@

In most practical manifold learning tasks data are often quite large, so we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud. This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest. Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is assumed to be large. To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_steps} (b), along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data. The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure. By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ from $I_t$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.6]{inconsistent_parameterization}
  \caption{Simulated data with $D = 2$. The curves shown in this plot represents embedding maps estimated by PME for each time point, with time indicated by the color of the curve. The $X$ and $Y$ axes represent the coordinates of the curve in $D$-dimensional space, while $R$ represents the estimated parameters associated with these coordinates. Randomness in the initialization of the PME algorithm may result in misaligned parameterizations of the estimated embedding maps. For most functions, the $X$ values increase with increasing values for $R$. However, two functions have $X$ values that decrease with increasing $R$. The coefficients for these two functions deviate enough from the other function coefficients that smoothing over the functions with respect to time becomes much more difficult.}
  \label{fig:inconsistent_parameterization}
\end{figure}

@@

Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}. Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation. Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings. The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin. Finally, the data were augmented with spherical coordinates in a manner similar to that described in Section \ref{s:LPME} to avoid the need to fit to a self-intersecting manifold.

@@

Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figures \ref{fig:adni_result}. Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit well to the data. There are two main sources of discrepancies between the data and the LPME estimate. First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation. The second difference between the LPME estimates and the observed data can be seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of the curves seen in the observed data.

Figure \ref{fig:adni_cross_section} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point. These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points. This figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each of the plots. Additionally, we see that, along the upper-right edge of the surface, there is a spot where the estimated manifold intersects with itself. This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME.

@@

Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:adni_lthal_cross_section}. While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:adni_lthal_cross_section} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
  \caption{Left Hippocampus Cross Section}
  \label{fig:adni_cross_section}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{adni_plots/adni_lthal_cross_section.png}
    \caption{Left Thalamus Cross Section}
    \label{fig:adni_lthal_cross_section}
\end{figure}

@@

In this article, we propose an extension to the Principal Manifold Estimation algorithm introduced by \cite{mengPrincipalManifoldEstimation2021}, which may be used to model longitudinal changes in a low-dimensional manifold underlying high-dimensional data. To the best of our knowledge, this is the first time a nonlinear manifold learning method has been adapted to a longitudinal setting. In simulated datasets where spurious changes in the structure to be estimated were introduced between time points, the LPME algorithm showed improved performance in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve algorithms at each time point. This improvement was seen across several manifold types. We also suggest a data augmentation approach that may allow for the circumvention of the inability to fit principal curve-based methods to some self-intersecting manifolds.

It is key to note here that the MSD values shown in these tables are comparing the various models' projections onto the estimated manifold to the true values underlying the data the models are fit on. This means that a model that performs well using this metric may not perform well when comparing the data to the model's projections onto the manifold. In fact, if there is a high level of inter-observation noise, then we should expect an estimated manifold that closely follows the true underlying manifold to show a relatively poor fit to the data at any given time point. This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are trying to estimate a curve or surface using the data available at a given time point, which may have noise introduced that increases the distance to the true manifold. When comparing the fit of the PME and PC methods to the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.

While the LPME algorithm demonstrated strong performance in a number of settings, it also has several limitations. First, because the method explicitly builds on the results of an estimate provided by the PME algorithm, the LPME algorithm will only fit properly in situations where PME fits properly. Because the output of PME is highly dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature. The option to use alternative methods, such as Locally Linear Embedding or Laplacian Eigenmaps, to generate the initial parameterization for the PME algorithm may help to alleviate this concern. There is also room to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise present.

@@

Another limitation stems from the performance of the LPME method on the hippocampus surfaces obtained from the ADNI dataset. It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus. This shortcoming could potentially be addressed in several ways. First, as discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated. Second, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated. Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}.

Additional work will be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities. The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods. However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin. The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations. While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.

>>>>

\usepackage[
backend=biber,
style=authoryear,
url=false,
isbn=false,
date=year,
doi=false,
pmid=false,
urldate=false
]{biblatex}

@@

\begin{figure}[ht]
  \centering
  \subfloat[\centering Data]{{\label{fig:lpme_step1} \includegraphics[width=8cm]{sim_case1_plot2}}}%
  \hfill
  \subfloat[\centering Step 1: Sample Size Reduction]{{\label{fig:lpme_step2} \includegraphics[width=8cm]{lpme_data_reduction}}}
  \vfill
  \subfloat[\centering Step 2: Initialization]{{\label{fig:lpme_step3} \includegraphics[width=8cm]{lpme_initialization}}}
  \hfill
  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\label{fig:lpme_step4} \includegraphics[width=8cm]{lpme_fitting}}}
  \caption{LPME Algorithm Steps. In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface. }
  \label{fig:lpme_steps}
\end{figure}

@@

In most practical manifold learning tasks data are often quite large, so we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud. This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest. Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is assumed to be large. To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data. The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure. By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ from $I_t$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

@@

\begin{figure}[h]
  \centering
  \includegraphics[height=7cm]{inconsistent_parameterization}
  \caption{Simulated data with $D = 2$. The curves shown in this plot represents embedding maps estimated by PME for each time point, with time indicated by the color of the curve. The $X$ and $Y$ axes represent the coordinates of the curve in $D$-dimensional space, while $R$ represents the estimated parameters associated with these coordinates. Randomness in the initialization of the PME algorithm may result in misaligned parameterizations of the estimated embedding maps. For most functions, the $X$ values increase with increasing values for $R$. However, two functions have $X$ values that decrease with increasing $R$. The coefficients for these two functions deviate enough from the other function coefficients that smoothing over the functions with respect to time becomes much more difficult.}
  \label{fig:inconsistent_parameterization}
\end{figure}

@@

Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}. Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation. Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings. The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in each dimension of Euclidean space and centered around the origin. Finally, the data were augmented with spherical coordinates in a manner similar to that described in Section \ref{s:LPME} to avoid the need to fit to a self-intersecting manifold.

@@

Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figure \ref{fig:adni_result}. Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit reasonably to the data. There are two main sources of discrepancies between the data and the LPME estimate. First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points. The second difference between the LPME estimates and the observed data can be seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of the curves seen in the observed data.

Figure \ref{fig:lhipp_cross_section} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point. These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points. This figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each of the plots. Additionally, we see that, along the upper-right edge of the surface, there is a spot where the estimated manifold intersects with itself. This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME. Notably, it also appears that the boundaries of the LPME-estimated surface tend to fall inside the boundaries of the PME-estimated surface, particularly at locations with the highest levels of curvature.

@@

To understand the impacts of these discrepancies on the estimated volume of the regions, the volume of the hippocampus was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point. The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume. The interior identification approach developed by \cite{mengPrincipalManifoldEstimation2021} is used to determine whether each voxel may be considered inside the surface of the structure or not. The code used to implement this approach is available at \texttt{https://github.com/rjzielinski/lpme-project}. The volume estimates can be found in Figure \ref{fig:lhipp_volumes}. Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.

Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:adni_lthal_cross_section}. While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:adni_lthal_cross_section} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus. This is reflected in the smaller discrepancies between volume estimates, as seen in Figure \ref{fig:lthal_volumes}, though visible variations between time points remain.

@@

\begin{figure}[h]
  \centering
  \subfloat[\centering Left Hippocampus Cross Section]{\label{fig:lhipp_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_cross_section}}%
  \hfill
  \subfloat[\centering Estimated Volumes]{\label{fig:lhipp_volumes} \includegraphics[width=8.5cm]{adni_plots/lhipp_volume}}
  \caption{Left Hippocampus Cross Section and Volume Estimates. In (a), we can see that the LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of surface in the observed data. In the fifth observation with $t=3.04$, the LPME-estimated surface has a gap in it, corresponding to an unexpectedly high volume estimate seen in (b). Volume estimates derived from the observed data are consistently higher than those obtained from the embedding functions of PME and LPME estimates, respectively.}
  \label{fig:adni_cross_section}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \subfloat[\centering Left Thalamus Cross Section]{\label{fig:lthal_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_lthal_cross_section}}%
  \hfill
  \subfloat[\centering Estimated Volumes]{\label{fig:lthal_volumes} \includegraphics[width=8.5cm]{adni_plots/lthal_volume}}
  \caption{Left Thalamus Cross Section and Volume Estimates. In (a), the cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface. The volume estimates in (b) show similar general trends to the hippocampus estimates, with the estimates reached using the observed data being higher than the PME- and LPME-based estimates. However, the differences between the estimates taken from the PME and LPME results are more comparable.}
  \label{fig:adni_lthal_cross_section}
\end{figure}

@@

In this article, we propose an extension to the Principal Manifold Estimation algorithm introduced by \cite{mengPrincipalManifoldEstimation2021}, which may be used to model longitudinal changes in a low-dimensional manifold underlying high-dimensional data. To the best of our knowledge, this is the first time a nonlinear manifold learning method has been adapted to a general-purpose longitudinal setting. We also suggest a data augmentation approach that circumvents the inability to fit principal curve-based methods to self-intersecting manifolds in select settings. In simulated datasets where spurious changes in the structure to be estimated were introduced between time points, the LPME algorithm showed improved performance in recovering the underlying manifold when compared to a naive approach of applying either the PME or principal curve algorithms at each time point. This improvement was seen across several manifold types.

It is key to note here that the MSD values shown in these tables compare the various models' projections onto the estimated manifold to the true values underlying the data the models are fit on. This means that a model that performs well using this metric may not perform well when comparing the model's projections onto the manifold with the data. In fact, if there is a high level of inter-observation noise, then we should expect an estimated manifold that closely follows the true underlying manifold to show a relatively poor fit to the data at any given time point. This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are trying to estimate a curve or surface using the data available at a given time point, which may have noise introduced that increases the distance to the true manifold. When comparing the fit of the PME and PC methods to the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.

While the LPME algorithm demonstrated strong performance in a number of settings, it also has several limitations. First, because the method explicitly builds on the results of an estimate provided by the PME algorithm, the LPME algorithm will only fit properly in situations where PME fits properly. Because the output of PME is dependent on the success of its initial parameterization, in practice this limits the success of LPME to situations where Isomap performs well, which excludes nonconvex manifolds, manifolds that have holes, or manifolds that have too high a level of curvature. The option to use alternative methods, such as Locally Linear Embedding or Laplacian Eigenmaps, to generate the initial parameterization for the PME algorithm may help to alleviate this concern. Opportunities also exist to improve the tuning of the initialization method, which may result in better performance, especially when considering data with different levels of noise present.

@@

Another limitation stems from the performance of the LPME method on the hippocampus surfaces obtained from the ADNI dataset. It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus. This shortcoming could potentially be addressed in several ways. First, as discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated. Second, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated. Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}. The volume estimates obtained from the outputs of the PME and LPME algorithms reflected the insufficient fit of these algorithms to the hippocampus data. However, the volume estimation process includes ad hoc approaches to address missing data, and the high level of variability in the estimated volumes of the thalamus despite visual indications of improved fit to this data suggests that this approach may itself be contributing to this variability. Further testing of this approach will be needed before the LPME algorithm will be viable for approximating the volume of brain regions.

Additional work will also be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities. The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods. However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to the hippocampus or more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin. The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations. While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.
